### 2022-10-18 chapters/benchmarking/benchmarking.tex

\chapter{Benchmarking Protocol to Measure Software Energy Consumption}\label{chapter:benchmarking}
% \section{Introduction \note{missing} }
This chapter covers ways to overcome empirical analysis challenges in energy consumption studies.
First, we go through the three components of a successful benchmark when performing energy-related experiments.
Section~\ref{sec:benchmarking_reproducibility} focuses on the "reproducibility" challenge---to deliver reproducible experiments without interfering with the energy measurement---while Section~\ref{sec:taming-the-energy-variation} discusses accuracy of software energy consumption.

>>>>

\chapter{Benchmarking Protocol to Measure Software Energy Consumption}\label{chapter:benchmarking}
% \section{Introduction \note{missing} }
This chapter covers ways to overcome empirical analysis challenges in energy consumption studies.
First, we go through the three components of a successful benchmark when performing energy-related experiments.
Section~\ref{sec:benchmarking_reproducibility} focuses on the "reproducibility" challenge---to deliver reproducible experiments without interfering with the energy measurement---while Section~\ref{sec:taming-the-energy-variation} discusses the \emph{accuracy} of software energy consumption.

### 2022-10-18 chapters/benchmarking/draft.tex
update : abstract ( en+fr)

Empirical measurements are critical to capture the effect of developers' choices on software energy consumption.
To accomplish this, one should not overlook the benchmarking pitfalls highlighted in the state of the art~\cite{van_der_kouwe_benchmarking_2018}.

Second, one should not ignore tremendous progress in computer science, which has led to a rise in the number of obsolete results.
Furthermore, when it comes to comparative research, the execution environment may impact the study itself.
Finally, between the exploratory experiment and the publication of the results, new candidates may have emerged, and others may have changed.

@@

\subsection{Docker Vs. Virtual Machine}
Despite the fact that Type 1 is more performant than Type 2, the latter is the most used in research, as most researchers tend to conduct their experiments on their own machines. Docker, on the other hand, is the most well-known container technology. In our case, we are more likely to promote Docker for two reasons:
\begin{enumerate}
    \item As previously stated in literature~\cite{van2016power,morabito_power_2015}, we require a lightweight orchestrator to limit the overhead on the energy usage of our experiments
    \item We need to communicate with the host OS because we are using hardware sensors to measure energy consumption.
\end{enumerate}

@@

\subsection{Extension}
% TODO : proofread
\subsubsection{Definition of Extention}
With the rapid evolution of the software industry, Even ensuring the reproducibility of the same research will not be enough.
Each day, new software versions are released, and new features are added. Even more, the goal of the research might evolve. As a result, nowadays, especially in comparative studies, it is essential to leave room for expansion.

@@

The core design of the experiment is responsible for running the experiment regardless of the context.
In the experiment, it is the only component that is allowed to interact with the SUT.
The orchestrator provides three interfaces:
\begin{itemize}
    \item \emph{Workload interface}: provides a set of functions that the workload should implement to be called by each candidate in the experiment. This interface is responsible for extending the experiment with new benchmarks.
    \item \emph{Observer interface}: provides a set of metrics that the orchestrator collects during the experiment. Implementing the observer interface allows the user to extend the experiment with new metrics.
    \item \emph{The Candidate interface}: It provides a set of functions that the candidate should implement to called  by the orchestrator. This interface is responsible for extending the experiment with new solutions.
\end{itemize}

>>>>

Empirical measurements are critical to capture the effect of developers' choices on software energy consumption.
To accomplish this, one should not overlook the benchmarking pitfalls highlighted in the state-of-the-art~\cite {van_der_kouwe_benchmarking_2018}.

Second, one should not ignore the tremendous progress in computer science, which has led to a rise in the number of obsolete results.
Furthermore, when it comes to comparative research, the execution environment may impact the study itself.
Finally, between the exploratory experiment and the publication of the results, new candidates may have emerged, and others may have changed.

@@

\subsection{Docker Vs. Virtual Machine}
Even though Type 1 is more performant than Type 2, the latter is the most used in research, as most researchers tend to conduct their experiments on their machines. Docker, on the other hand, is the most well-known container technology. In our case, we are more likely to promote Docker for two reasons:
\begin{enumerate}
    \item As previously stated in literature~\cite{van2016power,morabito_power_2015}, we require a lightweight orchestrator to limit the overhead on the energy usage of our experiments
    \item We need to communicate with the host OS because we are using hardware sensors to measure energy consumption.
\end{enumerate}

@@

\subsubsection{Definition of Extention}
With the rapid evolution of the software industry, Even ensuring the reproducibility of the same research will not be enough.
Each day, new software versions are released, and new features are added. Even more, the goal of the research might evolve. As a result, nowadays, especially in comparative studies, it is essential to leave room for expansion.

@@

The core design of the experiment is responsible for running the experiment regardless of the context.
In the experiment, it is the only component that is allowed to interact with the SUT.
The orchestrator provides three interfaces:
\begin{itemize}
    \item \emph{Workload interface}: provides a set of functions that the workload should implement to be called by each candidate in the experiment. This interface is responsible for extending the experiment with new benchmarks.
    \item \emph{Observer interface}: provides a set of metrics that the orchestrator collects during the experiment. Implementing the observer interface allows the user to extend the experiment with new metrics.
    \item \emph{The Candidate interface}: It provides a set of functions that the candidate should implement to be called by the orchestrator. This interface is responsible for extending the experiment with new solutions.
\end{itemize}

### 2022-10-18 chapters/benchmarking/experiment.tex

Every experiment is conducted on $100$ iterations, on multiple nodes and using the 3 NPB benchmarks we mentioned, with a warmup phase of 10 iterations for each experiment.
In most cases, we sought to evaluate the \emph{Standard Deviation} (STD), which is the most representative factor of the energy variation.
While running our experiments, we tried not to fall into the most common benchmarking "crimes"~\cite{DBLP:journals/corr/abs-1801-02381}.
As we study the STD difference of measurements we observed from empirical experiments, we use the bootstrap method~\cite{efron2000bootstrap} to randomly build multiple subsets of data from the original dataset, and we draw the STD density of those sets, as illustrated in Figure~\ref{fig:docker}.
Given the space constraints, we report on aggregated results for nodes, benchmarks, and workloads.
However, the raw data we collected is available through the public repository we published.\footnote{\url{https://github.com/anonymous-data/Energy-Variation}}
We believe this can help to achieve better and more reliable comparisons.
We mainly consider three different workloads in our experiments: \textsf{single process}, 50\,\%, and 100\,\%, to cover the low, medium, and high CPU usage when analyzing the effect of the studied parameters, respectively.
These workloads reflect the ratio of used PU count to the total available PU.

@@

\subsection{Docker \& Accuracy}
As the state of the art assesses the impact of Docker on energy consumption, one can also consider its impact on accuracy.
In other words:\\
\textbf{RQ:} does Docker affect the energy variation of the experiments?

To answer this question, we conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG} and \textsf{EP} in a Docker container and a flat binary format on three nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running the same compiled version of the benchmarks at 5\,\%, 50\,\% and 100\,\% workloads.
While Docker increases the energy consumption due to the extra layer it implements~\cite{eddie_antonio_santos_how}, it does not noticeably affect the energy variation.
The \emph{standard deviation} (STD) is even slightly smaller ($STD_{Docker}=192 mJ$,$STD_{Binary}=207 mJ$), taking into account the measurements errors and the OS activity.

@@

We can see the effect that has the \textsf{C-states off} mode when running a single-process application/benchmark.
The energy consumption varies 5~times less than the default mode.
In this case, only one CPU core is used among $2\times16$ physical cores.
The other cores are switched to a low consumption state when C-states are on. The switching operation causes a vital energy consumption difference between the cores and could be affected by other activities, such as the kernel activity, causing a notable energy consumption variation.
On the other hand, switching off the C-states would keep all the cores---even the unused ones---at a high-frequency usage.
This highly reduces the variation, but causes up to 50\,\% of extra energy consumption in this test ($Mean_{C-states-off}=11,665 mJ$,$Mean_{C-states-on}=7,641 mJ$).

@@

Even though the first and third strategies are very similar (both use hyper-threads, but only on one CPU for the first and two CPUs for the third), the variation gap between them is significant, as the variation in the first strategy is 30 times lower ($STD_{S1}=116 mJ$,$STD_{S3}=3,452 mJ$).
This shows that the hyper-threads technology is not the main reason behind the variation; the first strategy has even less variation than the second one and still uses the hyper-threading.

@@

The first way to do this is to evaluate the OS idle activity consumption and to compare it to a low workload running job.
Therefore, we ran 100 iterations of a single process benchmark \textsf{EP}, \textsf{LU} and \textsf{CG} on multiple nodes from the cluster \textsf{Dahu}, and compared the energy behavior of the node with its idle state on the same duration.
The aggregated results, illustrated in Figure~\ref{fig:os-idle}, depict that the idle energy variation is up to 140\,\% worst than when running a job, even if it consumes 120\,\% less energy ($Mean_{Job}=8,746 mJ$, $Mean_{Idle}=3,927 mJ$).
In fact, for the three nodes randomly picked from the cluster \textsf{Dahu}, the idle variation is way more important than when a test was running, even if it is a single process test on a $32$-cores node.
This result shows that OS idle consumption varies widely due to the lack of activity and the different CPU frequency states. However, it does not mean that this variation is responsible for the overall energy variation.
The OS behaves differently when a job is running, even if the amount of available cores is more than enough for the OS to keep its idle behavior when running a single process.

@@

\paragraph{OS Tuning}
An OS is a pack of running processes and services that might or not be required for its execution.
Even using a minimal version of Debian Linux, we could list many OS running services and processes that could be disabled/stopped without impacting the test execution.
This extra-tuning may not be the same, depending on the nature of the test or the OS.
Thus, we conducted a test with a deeply-tuned OS version.
We disabled all the services/processes that are not essential to the OS/test running, including the OS networking interfaces and logging modules. We only kept the strict minimum required for the experiment's execution.
Table~\ref{table:osmin} reports on the aggregated results for running single process measurements with the benchmarks \textsf{CG}, \textsf{LU} and \textsf{EP}, on three servers of the cluster \textsf{Dahu}, before and after tuning the OS.
Every cell contains the \emph{STD} value before the tuning, plus/minus a ratio of the energy variation after the tuning.
We notice that the energy variation varies less than 10\,\% after the extra-tuning.
We argue that this variation is not substantial, as it is not stable from one node to another.
Moreover, 10\,\% of the variation is not a representative difference due to many factors that can affect it as the CPU temperature or the measurement errors.

@@

Table~\ref{table:guidelines} gives a proper understanding of known factors, like the C-states and their variation reduction at low workloads.
However, it also lists some new factors we identified along the analysis we conducted in Section, such as the results related to the OS or the reboot mode.
Some guidelines are more useful/efficient for specific workloads, as shown in our experiments.
Thus, qualifying the workload before conducting the experiments can help choose the proper guidelines.
Other studied factors, such as Turbo~Boost or Speculative execution, are not mentioned in the guidelines due to the small effect observed in our study.

To validate the accuracy of our guidelines among a varied set of benchmarks on the one hand and their effect on the variation between identical machines, on the other hand, we ran seven experiments with benchmarks and real applications on a set of four identical nodes from the cluster \textsf{Dahu}, before (\textsf{normal} mode where everything is left to default and to the charge of the OS) and after (\textsf{optimized}) applying our guidelines.
Half of these experiments have been performed at a 50\,\% workload, and the other half on single process jobs.
The choice of these two workloads is related to the optimization guidelines that are mainly effective at low and medium workloads.
We note that we used the cluster \textsf{Dahu} over \textsf{Ecotype} to highlight the effect of the guidelines on the nodes where the variation is susceptible to be higher.

@@

\subsection{Threats to Validity}
Several issues affect the validity of our work.
For most of our experiments, we used the Intel RAPL tool, which has evolved along Intel CPU generations to be known as one of the most accurate tools for modern CPUs but still adds a significant overhead if we adopt a sampling at high frequency.
The other fine-grained measurement tool we used is \textsc{PowerAPI}.
It allows measuring the energy consumption at the granularity of a process or a Cgroup by dividing the RAPL global energy over the running processes using a power model.
The usage of \textsc{PowerAPI} adds an error margin because of the power model built over RAPL.
The RAPL tool mainly measures CPU and DRAM energy consumption.
However, even running CPU/RAM intensive benchmarks would keep a degree on uncertainty concerning the hard disk and networking energy consumption.
In addition, the operating system adds a layer of confusion and uncertainty.

>>>>

Every experiment is conducted on $100$ iterations, on multiple nodes, and using the 3 NPB benchmarks we mentioned, with a warmup phase of 10 iterations for each experiment.
In most cases, we sought to evaluate the \emph{Standard Deviation} (STD), which is the most representative factor of the energy variation.
While running our experiments, we tried not to fall into the most common benchmarking "crimes"~\cite{DBLP:journals/corr/abs-1801-02381}.
As we study the STD difference of measurements we observed from empirical experiments, we use the bootstrap method~\cite{efron2000bootstrap} to randomly build multiple subsets of data from the original dataset, and we draw the STD density of those sets, as illustrated in Figure~\ref{fig:docker}.
Given the space constraints, we report on aggregated results for nodes, benchmarks, and workloads.
However, the raw data we collected is available through the public repository we published.\footnote{\url{https://github.com/anonymous-data/Energy-Variation}}
We believe this can help to achieve better and more reliable comparisons.
We mainly consider three different workloads in our experiments: \textsf{single process}, 50\,\%, and 100\,\%, to cover the low, medium, and high CPU usage when analyzing the effect of the studied parameters, respectively.
These workloads reflect the ratio of used PU count to the total available PU.

@@

\subsection{Docker \& Accuracy}
As the state-of-the-art assesses the impact of Docker on energy consumption, one can also consider its impact on accuracy.
In other words:\\
\textbf{RQ:} does Docker affect the energy variation of the experiments?

To answer this question, we conducted a preliminary experiment by running the same benchmarks \textsf{LU}, \textsf{CG}, and \textsf{EP} in a Docker container and a flat binary format on three nodes of the cluster \textsf{Dahu} to assess if Docker induces an additional variation.
Figure~\ref{fig:docker} reports that this is not the case, as the energy consumption variation does not get noticeably affected by Docker while running the same compiled version of the benchmarks at 5\,\%, 50\,\%, and 100\,\% workloads.
While Docker increases the energy consumption due to the extra layer it implements~\cite{eddie_antonio_santos_how}, it does not noticeably affect the energy variation.
The \emph{standard deviation} (STD) is even slightly smaller ($STD_{Docker}=192 mJ$,$STD_{Binary}=207 mJ$), taking into account the measurements errors and the OS activity.

@@

We can see the effect that has the \textsf{C-states off} mode when running a single-process application/benchmark.
The energy consumption varies 5~times less than the default mode.
In this case, only one CPU core is used among $2\times16$ physical cores.
The other cores are switched to a low-consumption state when C-states are on. The switching operation causes a vital energy consumption difference between the cores and could be affected by other activities, such as the kernel activity, causing a notable energy consumption variation.
On the other hand, switching off the C-states would keep all the cores---even the unused ones---at a high-frequency usage.
This highly reduces the variation, but causes up to 50\,\% of extra energy consumption in this test ($Mean_{C-states-off}=11,665 mJ$,$Mean_{C-states-on}=7,641 mJ$).

@@

Even though the first and third strategies are very similar (both use hyper-threads, but only on one CPU for the first and two CPUs for the third), the variation gap between them is significant, as the variation in the first strategy is 30 times lower ($STD_{S1}=116 mJ$,$STD_{S3}=3,452 mJ$).
This shows that the hyper-threads technology is not the main reason behind the variation; the first strategy has even less variation than the second one and still uses hyper-threading.

@@

The first way to do this is to evaluate the OS idle activity consumption and compare it to a low-workload running job.
Therefore, we ran 100 iterations of a single process benchmark \textsf{EP}, \textsf{LU}, and \textsf{CG} on multiple nodes from the cluster \textsf{Dahu}, and compared the energy behavior of the node with its idle state on the same duration.
The aggregated results, illustrated in Figure~\ref{fig:os-idle}, depict that the idle energy variation is up to 140\,\% worst than when running a job, even if it consumes 120\,\% less energy ($Mean_{Job}=8,746 mJ$, $Mean_{Idle}=3,927 mJ$).
In fact, for the three nodes randomly picked from the cluster \textsf{Dahu}, the idle variation is way more important than when a test was running, even if it is a single process test on a $32$-cores node.
This result shows that OS idle consumption varies widely due to the lack of activity and the different CPU frequency states. However, it does not mean that this variation is responsible for the overall energy variation.
The OS behaves differently when a job is running, even if the amount of available cores is more than enough for the OS to keep its idle behavior when running a single process.

@@

\paragraph{OS Tuning}
An OS is a pack of running processes and services that might or not be required for its execution.
Even using a minimal version of Debian Linux, we could list many OS running services and processes that could be disabled/stopped without impacting the test execution.
This extra-tuning may not be the same, depending on the nature of the test or the OS.
Thus, we conducted a test with a deeply-tuned OS version.
We disabled all the services/processes that are not essential to the OS/test running, including the OS networking interfaces and logging modules. We only kept the strict minimum required for the experiment's execution.
Table~\ref{table:osmin} reports on the aggregated results for running single process measurements with the benchmarks \textsf{CG}, \textsf{LU} and \textsf{EP}, on three servers of the cluster \textsf{Dahu}, before and after tuning the OS.
Every cell contains the \emph{STD} value before the tuning, plus/minus a ratio of the energy variation after the tuning.
We notice that the energy variation varies less than 10\,\% after the extra-tuning.
We argue that this variation is not substantial, as it is not stable from one node to another.
Moreover, 10\,\% of the variation is not a representative difference due to many factors that can affect it as CPU temperature or measurement errors.

@@

Table~\ref{table:guidelines} gives a proper understanding of known factors, like the C-states and their variation reduction at low workloads.
However, it also lists some new factors we identified along the analysis we conducted in the previous Section, such as the results related to the OS or the reboot mode.
Some guidelines are more useful/efficient for specific workloads, as shown in our experiments.
Thus, qualifying the workload before conducting the experiments can help choose the proper guidelines.
Other studied factors, such as Turbo~Boost or Speculative execution, are not mentioned in the guidelines due to the small effect observed in our study.

To validate the accuracy of our guidelines among a varied set of benchmarks on the one hand and their effect on the variation between identical machines, on the other hand, we ran seven experiments with benchmarks and real applications on a set of four identical nodes from the cluster \textsf{Dahu}, before (\textsf{normal} mode where everything is left to default and the charge of the OS) and after (\textsf{optimized}) applying our guidelines.
Half of these experiments have been performed at a 50\,\% workload, and the other half on single process jobs.
The choice of these two workloads is related to the optimization guidelines that are mainly effective at low and medium workloads.
We note that we used the cluster \textsf{Dahu} over \textsf{Ecotype} to highlight the effect of the guidelines on the nodes where the variation is susceptible to be higher.

@@

\subsection{Threats to Validity}
Several issues affect the validity of our work.
For most of our experiments, we used the Intel RAPL tool, which has evolved along Intel CPU generations to be known as one of the most accurate tools for modern CPUs but still adds a significant overhead if we adopt a sampling at high frequency.
The other fine-grained measurement tool we used is \textsc{PowerAPI}.
It allows measuring the energy consumption at the granularity of a process or a Cgroup by dividing the RAPL global energy over the running processes using a power model.
The usage of \textsc{PowerAPI} adds an error margin because of the power model built over RAPL.
The RAPL tool mainly measures CPU and DRAM energy consumption.
However, even running CPU/RAM intensive benchmarks would keep a degree of uncertainty concerning the hard disk and networking energy consumption.
In addition, the operating system adds a layer of confusion and uncertainty.

### 2022-10-18 chapters/literature/draft_citations.tex

\citeauthor{bukh1992art} wrote a book about computer performance analysis, where they discussed some familiar topics that are relevant to statistical analysis, such as null hypotheses, chi-squared tests, regression, discrete event simulation, Bayes' theorem, how and when to use them for experimental design, measurement, simulation, and modeling for computer systems.
The article~\cite{brown2018issues} provides an intellectual framework for understanding the pervasiveness of mistakes in the scientific discovery process and presents methodological, cultural, and system-level techniques for minimizing the frequency of often seen errors.

@@

Other monitor tools have been realized by the manufacturers, such as IBM Power executive~\cite{koomey2011growth}, which allows their customers to monitor the power consumption and thermal behavior of the of BladeCenter systems in the data center.

@@

Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
We can observe a considerable variation of the energy consumption, not only among homogeneous nodes but also at the scale of a single node, reaching up to $25\,\%$ in this example.

\begin{figure}%[!htb]
    \center{\includegraphics[width=.9\linewidth]{imgs/motivation}}
    \caption{CPU energy variation for the benchmark \textsf{CG}}\label{fig:motivation}
\end{figure}
%% state of the art 
Some researchers started investigating the hardware impact of the energy variation of power consumption.
As an example, one can cite~\cite{borkar_designing_2005,tschanz_adaptive_2002} who reported that the leading cause of the variation of the power consumption between different machines is due to the \textbf{CMOS} manufacturing process of transistors in a chip.
\cite{heinrich_predicting} described this variation as a set of parameters, such as CPU Frequency and the thermal effect.
\subsection{Studying Hardware Factors}
This variation has often been related to the manufacturing process~\cite{coles_comparing_2014}. However, it has also been a subject of many studies, considering several aspects that could impact and vary the energy consumption across executions and on different chips.
On the one hand, the correlation between processor temperature and energy consumption was one of the most explored paths.
Kistowski~\emph{et~al.} showed in~\cite{joakim_v_kisroski_variations_2016} that identical processors could exhibit significant energy consumption variation with no close correlation with the processor temperature and performance.
On the other hand, the authors of~\cite{wang_potential_2018} claimed that the processor thermal effect is one of the most contributing factors to the energy variation, and the CPU temperature and the energy consumption variation are tightly coupled (up to 16\% increase in the variation when the temperature changed from 37.7°C to 74.5°C ).

@@

As for the impact of the source code on the energy consumption, we can cite~\cite{pinto_comprehensive_2016,fernandes_assisting_2017} where the authors investigated the impact of Java collections on energy usage based on collection size and its usage (insertion, removal, search) and They provide some insights into the energy efficiency of specific collections under various scenarios.

@@

We begin with Aequitas~\cite{ribic2016aequitas}, a system that allows parallel applications to live on co-managed power domains (sharing the same CPU). The technology is founded on the premise that coexisting programs can regard power-management hardware as a shared resource and collaborate on power management decisions. As a result, it accomplishes its purpose by scheduling these applications with a time-slicing technique (for example, round robin). The authors claim that their strategy achieves a 12.9\% improvement while incurring only a 2.5\% performance cost.

@@

%  TODO : rephrase this to fit the order of the chapters 
This work aims to present a set of guidelines to create a benchmarking system to measure different programs' energy consumption.
After that, we will use this system to compare the energy consumption of different programming languages. We will extend the work of \citeauthor{pereira_energy_2017} to a closer distance to the production environment by comparing a set of use cases.

>>>>

\citeauthor{bukh1992art} wrote a book about computer performance analysis, where they discussed some familiar topics that are relevant to statistical analysis, such as null hypotheses, chi-squared tests, regression, discrete event simulation, Bayes' theorem, how and when to use them for experimental design, measurement, simulation, and modeling for computer systems.
The article~\cite{brown2018issues} provides an intellectual framework for understanding the pervasiveness of mistakes in the scientific discovery process and presents methodological, cultural, and system-level techniques for minimizing the frequency of often-seen errors.

@@

Other monitor tools have been realized by the manufacturers, such as IBM Power executive~\cite{koomey2011growth}, which allows their customers to monitor the power consumption and thermal behavior of the BladeCenter systems in the data center.

@@

Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
We can observe a considerable variation of energy consumption, not only among homogeneous nodes but also at the scale of a single node, reaching up to $25\,\%$ in this example.

\begin{figure}%[!htb]
    \center{\includegraphics[width=.9\linewidth]{imgs/motivation}}
    \caption{CPU energy variation for the benchmark \textsf{CG}}\label{fig:motivation}
\end{figure}
%% state of the art 
Some researchers started investigating the hardware impact of the energy variation of power consumption.
As an example, one can cite~\cite{borkar_designing_2005,tschanz_adaptive_2002} who reported that the leading cause of the variation of the power consumption between different machines is due to the \textbf{CMOS} manufacturing process of transistors in a chip.
\cite{heinrich_predicting} described this variation as a set of parameters, such as CPU Frequency and the thermal effect.
\subsection{Studying Hardware Factors}
This variation has often been related to the manufacturing process~\cite{coles_comparing_2014}. However, it has also been the subject of many studies, considering several aspects that could impact and vary the energy consumption across executions and on different chips.
On the one hand, the correlation between processor temperature and energy consumption was one of the most explored paths.
Kistowski~\emph{et~al.} showed in~\cite{joakim_v_kisroski_variations_2016} that identical processors could exhibit significant energy consumption variation with no close correlation with the processor temperature and performance.
On the other hand, the authors of~\cite{wang_potential_2018} claimed that the processor thermal effect is one of the most contributing factors to the energy variation, and the CPU temperature and the energy consumption variation are tightly coupled (up to 16\% increase in the variation when the temperature changed from 37.7°C to 74.5°C ).

@@

As for the impact of the source code on energy consumption, we can cite~\cite{pinto_comprehensive_2016,fernandes_assisting_2017} where the authors investigated the impact of Java collections on energy usage based on collection size and its usage (insertion, removal, search) and They provide some insights into the energy efficiency of specific collections under various scenarios.

@@

We begin with Aequitas~\cite{ribic2016aequitas}, a system that allows parallel applications to live on co-managed power domains (sharing the same CPU). The technology is founded on the premise that coexisting programs can regard power-management hardware as a shared resource and collaborate on power-management decisions. As a result, it accomplishes its purpose by scheduling these applications with a time-slicing technique (for example, round robin). The authors claim that their strategy achieves a 12.9\% improvement while incurring only a 2.5\% performance cost.

@@

This work aims to present a set of guidelines to create a benchmarking system to measure different programs' energy consumption.
Following that, we will investigate the energy consumption of two well-known programming languages: Java and Python.
Finally, we extend this analysis to cover various programming languages
In this study, We will extend the work of \citeauthor{pereira_energy_2017} to a closer distance to the production environment.

### 2022-10-18 chapters/optimization/python-dir/python_interperters.tex

\section{Python Interpreters}\label{python:sec_interpreters}
We used the official version of Python for the first studies because the goal was mainly to highlight the code structure's impact on energy consumption.
One main drawback of the previous method is the work to be done to update the existing code base to reduce energy consumption.
Despite the availability of numerous tools for code refactoring and automatic code optimization, the optimization process remains a manual effort. This is because code optimization is a task that requires extensive understanding of both the codebase and the language itself.

To avoid such hustle, we tried to find a non-intrusive approach to make the Python code more eco-friendly without altering its structure.
Python is an interpreted language, which led to many initiatives to implement their interpreter to improve one or many aspects of the Python code. 
Therefore, in this section, we will examine at how different Python interpreters affect the energy consumption of software, and in which ca
se, one should use a non-conventional interpreter to save the energy consumption of their application.

\subsection{Python runtimes}\label{python:sec_interpreters}

@@

            A JIT API for CPython, same purpose as Parakeet and Hope;

@@

            A source-to-source transpiler that translates the Python code into Go before being compiled to a binary executable.

@@

We did not consider other implementations because either the project aborted many years ago or it has insufficient Python features.
After the inventory of those implementations, we filtered them.
To keep only the versions that are still maintained and support most Python features.
We then classified them into three categories depending on their integration with the Python code.
In \Cref{fig:pythonimplementations}, we describe the implementations we kept, each implementation's version, and its category.

@@

\subsection{Results \& Findings}
This part will be dedicated to the results of the experiments and statistical analysis of these results.
First and foremost, we want to know whether our samples have a normal distribution. As a result, we ran a Shapiro-Wilk~\cite{shapiro1968comparative} test on our data.
Even though some implementations had a $p$-value higher than $a=0.05$ (such as CPython and ActivePython), other implementations like PyPy and Numba presented a $p$-values smaller than 0.01, which leads to rejecting the hypothesis $H0$ that all the distributions follow a normal distribution.
Therefore, to compare the different implementations, we performed the non-parametric Mann-Whitney U test~\cite{zimmerman1987comparative}, with CPython2 and CPython3 as base references.

@@

Except for GraalPython, one can notice three main clusters.
The first cluster contains the reference implementations (CPython2 and CPython3) and the implementations that are based on interpreters (IronPython, IntelPython, ActivePython, Jython, and Nuitka).
Moreover, each implementation is the closest to its reference Python version.
The interpreters that are based on other virtual machines, such as Jython with JVM and IronPython with .NET, behave slightly differently from the reference implementation.
As for the MicroPython, It is the furthest from the references due to his behavior toward exceptions.
The second cluster groups the interpreter implementations that are based on a JIT compiler (PyPy2 and PyPy3).
Closer to them, we find Shedskin, which is a translator that converts the Python code into C++ code to be later compiled into binary code.
Table~\ref{table:manwithyu} shows that this cluster is by far the most energy-efficient one.
While Numba2 and Numba3 are also JIT libraries, they differ from PyPy due to their manual optimization.
Unlike PyPy, which decides the code parts to optimize, Numba gives this responsibility to the developer.
\bigskip

@@

\subsection{Conclusion}
One can conclude that the Python interpreter used considerably impacts the programs' energy consumption.
Furthermore, the absence of a universal solution makes this investigation more intriguing.
While most of the candidates employed \textsf{Python2}, \textsf{Python3} provided better compatibility coverage.
Except for \textsf{Pypy}, a standout interpreter, the optimization strategy substantially impacted energy usage because each class was grouped regardless of the Python version.
Finally, JIT, a technology that converts code into machine code at runtime, was responsible for the significant gain in energy efficiency.
The usage of translators, such as \textsf{Shedskin} confirmed the effect of compiling Python code into machine code. Hence, \textsf{PyPy}, an interpreter with a built-in JIT, is within the same class as the complied ones, such as \textsf{Shedskin}.

>>>>

\section{Python Interpreters}\label{python:sec_interpreters}
We used the official version of Python for the first studies because the goal was mainly to highlight the code structure's impact on energy consumption.
One main drawback of the previous method is the work to be done to update the existing code base to reduce energy consumption.
Despite the availability of numerous tools for code refactoring and automatic code optimization, the optimization process remains a manual effort. This is because code optimization is a task that requires an extensive understanding of both the codebase and the language itself.

To avoid a such hustle, we tried to find a non-intrusive approach to make the Python code more eco-friendly without altering its structure.
Python is an interpreted language, which led to many initiatives to implement their interpreter to improve one or many aspects of the Python code. 
Therefore, in this section, we will examine how different Python interpreters affect the energy consumption of software, and in which case, one should use a non-conventional interpreter to save the energy consumption of their application.

\subsection{Python Runtimes}\label{python:sec_interpreters}

@@

            A JIT API for CPython, the same purpose as Parakeet and Hope;

@@

            A source-to-source transpiler that translates the Python code into Go before being compiled into a binary executable.

@@

We did not consider other implementations because either the project was aborted many years ago or it has insufficient Python features.
After the inventory of those implementations, we filtered them.
To keep only the versions that are still maintained and support most Python features.
We then classified them into three categories depending on their integration with the Python code.
In \Cref{fig:pythonimplementations}, we describe the implementations we kept, each implementation's version, and its category.

@@

\subsection{Results \& Findings}
This part will be dedicated to the results of the experiments and statistical analysis of these results.
First and foremost, we want to know whether our samples have a normal distribution. As a result, we ran a Shapiro-Wilk~\cite{shapiro1968comparative} test on our data.
Even though some implementations had a $p$-value higher than $a=0.05$ (such as CPython and ActivePython), other implementations like PyPy and Numba presented $p$-values smaller than 0.01, which leads to rejecting the hypothesis $H0$ that all the distributions follow a normal distribution.
Therefore, to compare the different implementations, we performed the non-parametric Mann-Whitney U test~\cite{zimmerman1987comparative}, with CPython2 and CPython3 as base references.

@@

Except for GraalPython, one can notice three main clusters.
The first cluster contains the reference implementations (CPython2 and CPython3) and the implementations that are based on interpreters (IronPython, IntelPython, ActivePython, Jython, and Nuitka).
Moreover, each implementation is the closest to its reference Python version.
The interpreters that are based on other virtual machines, such as Jython with JVM and IronPython with .NET, behave slightly differently from the reference implementation.
As for the MicroPython, It is the furthest from the references due to his behavior toward exceptions.
The second cluster groups the interpreter implementations that are based on a JIT compiler (PyPy2 and PyPy3).
Closer to them, we find Shedskin, which is a translator that converts Python code into C++ code to be later compiled into binary code.
Table~\ref{table:manwithyu} shows that this cluster is by far the most energy-efficient one.
While Numba2 and Numba3 are also JIT libraries, they differ from PyPy due to their manual optimization.
Unlike PyPy, which decides the code parts to optimize, Numba gives this responsibility to the developer.
\bigskip

@@

\subsection{Summary}
One can conclude that the Python interpreter used considerably impacts the programs' energy consumption.
Furthermore, the absence of a universal solution makes this investigation more intriguing.
While most of the candidates employed \textsf{Python2}, \textsf{Python3} provided better compatibility coverage.
Except for \textsf{Pypy}, a standout interpreter, the optimization strategy substantially impacted energy usage because each class was grouped regardless of the Python version.
Finally, JIT, a technology that converts code into machine code at runtime, was responsible for the significant gain in energy efficiency.
The usage of translators, such as \textsf{Shedskin} confirmed the effect of compiling Python code into machine code. Hence, \textsf{PyPy}, an interpreter with a built-in JIT, is within the same class as the complied ones, such as \textsf{Shedskin}.
